---
title: "PA1_template"
author: "Eric Tchepannou"
date: "19 July 2015"
output:
  html_document:
    fig_caption: yes
    highlight: tango
    keep_md: yes
    theme: journal
    toc: yes
---

## Loading and preprocessing the data

```{r}
library(dplyr)
library(ggplot2)

data <- read.csv(file = "activity.csv", sep = ",")
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

### Calculate the total number of steps taken per day

```{r}
total_steps_per_day <- aggregate(data$steps, by=list(data$date), FUN=sum)
str(total_steps_per_day)
head(total_steps_per_day)
```

### If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r}
g <- ggplot(total_steps_per_day, aes(x = Group.1, y = x)) +
        geom_histogram(stat = "identity") +
        labs(x = "Interval",
             y = "Total number of steps",
             title = "Total number of steps taken each day")


print(g);
```

### Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean_steps_per_day <- aggregate(data$steps, by=list(data$date), FUN=mean, na.rm = T)
str(mean_steps_per_day)
head(mean_steps_per_day)

median_steps_per_day <- aggregate(data$steps, by=list(data$date), FUN=median, na.rm=T)
str(median_steps_per_day)
head(median_steps_per_day)
```

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
mean_steps_per_intv <- aggregate(data$steps, by=list(data$interval), FUN=mean, na.rm = T)

g <- ggplot(mean_steps_per_intv, aes(x = Group.1, y = x)) +
        geom_line() +
        labs(x = "Interval",
             y = "Average number of steps taken",
             title = "Average daily activity pattern")

print(g)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r comment="The 5-minute interval with the maximum number of steps"}
index <- which(mean_steps_per_intv$x == max(mean_steps_per_intv$x))
mean_steps_per_intv$Group.1[index]
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
length(data[is.na(data)])
```

### Use the mean to replace NAs
```{r}
mean_steps_per_intv <- aggregate(data$steps, by=list(data$interval), FUN=mean, na.rm = T)
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in.
``` {r}
new_data <- mutate(data, new_steps = ifelse(is.na(data$steps), mean_steps_per_intv$x, data$steps))
str(new_data)
head(new_data)
```

### Make a histogram of the total number of steps taken each day
``` {r}
new_total_steps_per_day <- aggregate(new_data$new_steps, by = list(new_data$date), FUN = sum, na.rm = T)
g <- ggplot(new_total_steps_per_day, aes(x = Group.1, y = x)) +
        geom_histogram(stat = "identity") +        
        labs(x = "Interval",
             y = "Total number of steps",
             title = "Total number of steps taken each day")


print(g);
```

Calculate and report the mean and median total number of steps taken per day

```{r}
new_mean_steps_per_day <- aggregate(new_data$new_steps, by = list(new_data$date), FUN = mean, na.rm = T)
str(new_mean_steps_per_day)
head(new_mean_steps_per_day)
new_median_steps_per_day <- aggregate(new_data$new_steps, by = list(new_data$date), FUN = median, na.rm = T)
str(new_median_steps_per_day)
head(new_median_steps_per_day)
```

The value differ from the estimates from the first part of the assignment. Imputing missing data on the estimates will increase the total daily number of steps taken.

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
day_data <- mutate(new_data, day = ifelse(weekdays(as.Date(new_data$date), T) == "Sat" | weekdays(as.Date(new_data$date), T) == "Sun", "weekend", "weekday"))
day_data$day <- as.factor(day_data$day)
str(day_data)
head(day_data)
```

```{r}
agg_day_data <- aggregate(day_data$steps, by=list(day_data$interval, day_data$day), mean, na.rm = T)

g <- ggplot(agg_day_data, aes(x = Group.1, y = x)) +
        geom_line() +
        geom_smooth(method = "lm") +
        facet_grid(facets = . ~ Group.2) +
        labs(x = "5-minute interval",
             y = "Average number of steps taken",
             title="Activity patterns between weekdays and weekends")

print(g)

```